Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-26 paper/main.tex
update figures and manuscript

Let $S$ represent the true (but unknown) underlying power matter spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence
(as visualized in Figure \ref{fig:plot_data}).  We thus specify a dense 
covariance matrix $\Sigma_\varepsilon$ that simulataneously
allows us to model the correlated errors and build domain-matter expertise
into the prior. To this end, we embrace variations of the Mat\'ern kernel 
\citep{stein1999interpolation}, tailored to each cosmology (more on this in later 
subsections).

@@

We compare DGP.FCO to competing models with a simulation study where the underlying 
true curve is known. Following \cite{moran2024dpc}, we consider two test functions 
defined as
\[
\begin{aligned}
  f_1(x) &= m_1 \exp\left(-\frac{u_1x}{2}\right) * 
  \cos\left(x\sqrt{25-\left(\frac{u_1}{2}\right)^2}\right) - \frac{m_1x}{5}\\
  f_2(x) &= \exp\left(-m_2(x-3)^2\right) + 
  \exp\left(-u_2(x-1)^2\right)-0.05\sin\left(8(x-1.9)\right).
\end{aligned}
\]
For additional variation, the values of $m_1, m_2, u_1,$ and $u_2$ are randomly selected from
$m_1 \sim \text{Uniform}(0.5,1.5)$, $u_1 \sim \text{Uniform}(1.5,2.5)$, and $m_2,u_2 
\sim \text{Uniform}(0.6,1.4)$. The outputs of $f_1$ and $f_2$ (conditional on $m_1/u_1$ and 
$m_2,u_2$), determine the true function values for each simulation.

@@

where $\boldsymbol{\gamma}_i = [\gamma_i^{(1)}, \gamma_i^{(2)}, \ldots, \gamma_i^{(m)}]^\top$ is the vector of 
$m$ observed PC weights for the $i$th component, $R$ is the $m \times m$ correlation matrix between 
all pairs of training inputs defined above, and $r(\psi^*) \in \mathbb{R}^m$ is the vector of correlations between 
the test input $\psi^*$ and each training input $\psi_j$. To illustrate, in the right panel of Figure \ref{fig:mean_PCs_oneW}, 
we show how the predicted GP mean for the weight of the fifth PC ($\gamma_5$) for Mira-Titan 
varies as a function of the first cosmological parameter $\psi_1$, while the other parameters are fixed at a (scaled)
midpoint value of 0.5.

@@

\textcolor{red}{(@Steve - we removed the text earlier that talked about the training/testing split in the
wrong section.  You need to introduce it here now.)} From our $m=32$ posterior means using DGP.FCO, 
these are input into SEPIA \textcolor{red}{(What does this mean?  Just refer to the equations from the previous
section for how we obtain predictions.)} in order to predict power spectra
for the $N=32$ test cosmologies.

Since CAMB offers the ``true'' infinite-resolution spectra, we have the opportunity to benchmark our
results in two ways.  First, we may compare our predictions to the true power spectrum for each held-out
cosmology.  Second, we may leverage the infinite-resolution spectra for the training cosmologies in
place of our estimated $\hat{S}_j$, to get a sense of the ``best-case" predictions that could originate
from the PC decomposition.  Figure \ref{fig:pca_preds_v_camb} compares our predictions (DGP.FCO + PC, red)
to this ``best-case'' prediction (true + PC, blue) and the actual true spectrum for \textcolor{red}{10?} 
of the held-out test cosmologies.  There is a high level of agreement between the pairs of predictions 
across cosmologies.  We then compare the mean squared errors of our predictions and the ``best-case'' 
predictions across all held-out cosmologies, with results shown in Figure \ref{fig:mse_camb}.  We again
see only minimal differences, which we take to indicate that DGP.FCO is offering effective
estimates of the true spectra for each cosmology.

\begin{figure}
    \centering
    \includegraphics[width=4in]{preds_diff1_4_8.jpg}
    \caption{Predictions of power matter spectra for 16 of the CAMB test cosmologies,
             after centering and scaling. 
             The true infinite resolution run is shown in black, the prediction 
             (trained on the infinite resolution) is in blue, and the prediction from 
             the DGP.FCO posterior mean is in red.
             \textcolor{red}{(This figure does not look polished.  Make it wider.  I recommend
             showing 2 rows of 5 plots.  Also, change line types so we are less reliant on color.
             Missing x-axis label.  No need to number the panels.)}}   
    \label{fig:pca_preds_v_camb}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=3in]{mse_dot.jpg}
    \caption{Comparison of MSE for CAMB predictions, when training with DGP.FCO posterior 
             means (left), and training with the ``true" infinite-resolution spectra (right).
             \textcolor{red}{(This figure also needs refinement.)}}   
    \label{fig:mse_camb}
\end{figure}

@@

Here we show prediction results for $N=6$ held-out cosmologies for the Mira-Titan dataset. 
\textcolor{red}{(This sentence is misleading.... we only offer one figure that shows RMSE.)}
We compare our method with Cosmic Emu \citep{moran2023mira}, the state-of-the-art emulator 
constructed on the same $m=111$ Mira-Titan cosmologies. Cosmic Emu uses a Bayesian approach 
where the ``true'' spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep{moran2024dpc}.  \textcolor{red}{(Add sentence about hwo Cosmic Emu also uses a PC
approach to predict power matter spectra for held-out cosmologies as a function of $\psi$.)}

%Both Cosmic Emu and DGP.FCO are capable of modeling a true underlying functions based 
%upon multiple functional realizations, but our approach utilizes DGPs within a 
%hierarchical model in lieu of deep process convolutions. After obtaining the estimated 
%smooth spectra from the training cosmologies, each method utilizes GPs in order to 
%estimate the weights of the PCs from the estimated spectra. Below, we compare prediction 
%performance between Cosmic Emu and DGP.FCO for the $N=6$ held-out cosmologies. 
% \textcolor{magenta}{In addition, we can also predict over a large uniform or full 
% factorial design to obtain estimated main effects for each cosmological parameter in 
% $\psi$, interactions amongst the $p_\psi=8$ parameters, and decompose the overall 
% variation by main effects, two-way interactions, and so on. Should this be included?}

Using our proposed approach and Cosmic Emu, we obtain two sets of predictions for $\hat{S}_t(\psi_t)$ 
with $t\in\{1,\ldots,6\}$ indexing the six held-out cosmologies.  To summarize and compare
performance, we evaluate the root mean squared error across $t$ for each wavenumber $x_i = \log_{10}(k_i)$.
\textcolor{red}{(This is again missing key details.  What is the truth you are using to calculate
RMSE???)}  The left panel of Figure \ref{fig:plot_rmse_k} shows these RMSEs for the held-out cosmologies.
Comparisons of the spectrum estimation for training cosmologies are shown as well (using RMSE against 
the weighted average of observations). \textcolor{red}{(What does this previous sentence mean??? We
have discussed that the weighted average is not a good estimate of the true spectrum.  So we should not
be comparing our estimated means to the weighted averages.  I am so confused....)}

For predictions, DGP.FCO achieves the greatest 
reduction in RMSE in the region where perturbation theory and the low-resolution runs 
are deemed unbiased. For the region where only the high-resolution run is unbiased, 
Cosmic Emu has a lower average RMSE. Over all $k$ values and predictions, DGP.FCO has 
an average RMSE of 0.0089 and Cosmic Emu has an average RMSE of 0.0106. Our RMSE is 
lower at 56\% of $k$ values considered. The individual predictions for each method and 
cosmology are shown in Figure \ref{fig:plot_pred_1to6}.  \textcolor{red}{(Why is this figure in the appendix?
But even more concerning, why are we subtracting the weighted average?  In every other plot we subtract our
estimated mean.  This switch to using the weighted average in the final subsection is very concerning
to me.  Also - what is the reader supposed to be gleaning from each of these plots?  Any time you
have a figure, you need to include narration in the text that explains what the takeaways are.)}

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{rmse_by_k.png}
    \caption{{\it Left:} Average RMSE of each prediction method across all $k$ values. 
             {\it Right:} Average RMSE for each method when using the held-out computer 
             model runs to train the model. In each case, the truth is assumed to 
             be the weighted average of the computer model runs.
             \textcolor{red}{(Shouldn't the average be across the 6 cosmologies?  Saying it 
             is across the $k$ values is misleading.  Again, I am concerned for us to use
             the weighted average as the truth.  This seems to go starkly against things
             that we have said earlier in the paper.)}}
    \label{fig:plot_rmse_k}
\end{figure}

% \begin{figure}[ht]
%     \centering
%     \includegraphics[width=6in]{rmse_box.png}
%     \caption{Boxplot of all RMSEs of each method across all six test models. 
%.             From top to bottom: box plots for our training method, process 
%              convolution training method, Cosmic EMU predictions, and our DGP-PC predictions.}
%     \label{fig:plot_rmse_box}
% \end{figure}

@@

\textcolor{red}{(@Steve - I think you should revisit this section once the rest of the
revisions are done.  Make sure it aligns with what the paper ends up being.  I am not going
through this sectio myself at this time.)}

In this work, we propose a novel Bayesian hierarchical model which utilizes DGPs 
in order to estimate a smooth underlying function from correlated functional observations 
(as cosmological computer model runs). In addition to the compositional form of the DGP, 
we include a third component modeling the correlated errors for each spectrum. 
With all smooth spectra estimated from the training set of 111 batches of model runs, 
we can predict spectra for unobserved cosmologies.

Multiple avenues for future research emerge: in addition to estimating dark matter 
power spectra, batches of hydrodynamical simulations exist, and this method could 
easily be adapted to handle such datasets. Other applications outside of cosmology 
can be explored, wherever multiple functional outputs exist, perhaps with spatial 
correlation (for example, observing multiple tropical cyclone forecasts). More detailed 
sensitivity analyses for the cosmological parameters can be explored to gain insights 
into how each of these parameters interact, and which are most significant. From a 
modeling perspective, changes to the model could include modeling the latent layer 
$W$ jointly across different cosmologies, as well as considering different redshift 
values other than $z=0$. Leveraging a hybrid model between DGP.FCO and Cosmic Emu 
may also help to obtain accurate estimates on the lower and higher ends of $k$ values, 
respectively. Additionally, Bayesian smoothing spline ANOVA (BSS-ANOVA) or other methods 
could be compared in addition to DGPs.

@@

\section{Mira-Titan Predictive Plots}
\label{sec:apdx_mira_plots}

% \textcolor{magenta}{Steve The different appendices from dissertation as necessary. 
% Can also put details of dataset in supplementary material.} 
% \textcolor{red}{Fix references as necessary, e.g. Moran et al. 2024.}

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{pred_1to6.png}
    \caption{Results of all six predictions for both methods on the held-out 
    cosmologies (weighted average subtracted). CosmicEMU is dashed, DGP.FCO is dotted, 
    and solid gray lines show results of held-out computer model runs.}
    \label{fig:plot_pred_1to6}
\end{figure}

>>>>

Let $S$ represent the true (but unknown) underlying power matter spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2023mira}, we find it
essential to account for the smoothly varying spatial dependence
(as visualized in Figure \ref{fig:plot_data}).  We thus specify a dense 
covariance matrix $\Sigma_\varepsilon$ that simulataneously
allows us to model the correlated errors and build domain-matter expertise
into the prior. To this end, we embrace variations of the Mat\'ern kernel 
\citep{stein1999interpolation}, tailored to each cosmology (more on this in later 
subsections).

@@

We compare DGP.FCO to competing models with a simulation study where the underlying 
true curve is known. Following \cite{moran2023mira}, we consider two test functions 
defined as
\[
\begin{aligned}
  f_1(x) &= m_1 \exp\left(-\frac{u_1x}{2}\right) * 
  \cos\left(x\sqrt{25-\left(\frac{u_1}{2}\right)^2}\right) - \frac{m_1x}{5}\\
  f_2(x) &= \exp\left(-m_2(x-3)^2\right) + 
  \exp\left(-u_2(x-1)^2\right)-0.05\sin\left(8(x-1.9)\right).
\end{aligned}
\]
For additional variation, the values of $m_1, m_2, u_1,$ and $u_2$ are randomly selected from
$m_1 \sim \text{Uniform}(0.5,1.5)$, $u_1 \sim \text{Uniform}(1.5,2.5)$, and $m_2,u_2 
\sim \text{Uniform}(0.6,1.4)$. The outputs of $f_1$ and $f_2$ (conditional on $m_1/u_1$ and 
$m_2,u_2$), determine the true function values for each simulation.

@@

where $\boldsymbol{\gamma}_i = [\gamma_i^{(1)}, \gamma_i^{(2)}, \ldots, \gamma_i^{(m)}]^\top$ is the vector of 
$m$ observed PC weights for the $i$th component, $R$ is the $m \times m$ correlation matrix between 
all pairs of training inputs defined above, and $r(\psi^*) \in \mathbb{R}^m$ is the vector of correlations between 
the test input $\psi^*$ and each training input $\psi_j$. To illustrate, in the right panel of Figure \ref{fig:mean_PCs_oneW}, 
we show how the predicted GP mean for the weight of the fifth PC ($\gamma_5$) for Mira-Titan 
varies as a function of the first cosmological parameter $\psi_1$, while the other parameters are fixed at a (scaled)
midpoint value of 0.5. With these estimates $\hat{\gamma}(\psi^*)$, we predict power spectra at 
held-out cosmologies for both the CAMB and Mira-Titan datasets.

@@

In order to predict power spectra from the CAMB dataset, we split the 64 cosmologies such that $m=32$
are used to train the DGP.FCO model and the remaining $N=32$ are reserved as hold-out cosmologies for
prediction. With our $m=32$ posterior means using DGP.FCO, we obtain $N=32$ predicted power spectra from
the functional prediction model outlined in Section \ref{subsec:pca} for the test cosmologies.

\begin{figure}
    \centering
    \includegraphics[width=0.85\textwidth]{pred_diffs_CAMB.jpeg}
    \caption{Predictions of power matter spectra for each of the 32 CAMB test cosmologies,
             after centering and scaling (to accentuate differences). 
             The true infinite resolution run is shown in black, the prediction 
             trained on the infinite resolution spectra is in dashed blue, and the prediction from 
             the DGP.FCO posterior means is in dotted orange.}
    \label{fig:pca_preds_v_camb}
\end{figure}

Since CAMB offers the ``true'' infinite-resolution spectra, we have the opportunity to benchmark our
results in two ways.  First, we may compare our predictions to the true power spectrum for each held-out
cosmology.  Second, we may leverage the infinite-resolution spectra for the training cosmologies in
place of our estimated $\hat{S}_j$, to get a sense of the ``best-case" predictions that could originate
from the PC decomposition.  Figure \ref{fig:pca_preds_v_camb} compares our predictions (DGP.FCO + PC, orange)
to these ``best-case'' predictions (true + PC, blue) and the actual true spectra for each of the 32
held-out test cosmologies.  There is a high level of agreement between the pairs of predictions across 
cosmologies. 
We then compare the mean squared errors of our predictions and the ``best-case'' 
predictions across all held-out cosmologies, with results shown in Figure \ref{fig:mse_camb}.  We again
see only minimal differences, which we take to indicate that DGP.FCO is offering effective
estimates of the true spectra for each cosmology.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mse_dot.jpeg}
    \caption{Comparison of MSE for CAMB predictions, when training with DGP.FCO posterior 
             means and training with the ``true" infinite-resolution spectra. The identity line 
             provides equal MSE values amongst both methods.}   
    \label{fig:mse_camb}
\end{figure}

@@

Here we show prediction results for $N=6$ held-out cosmologies for the Mira-Titan dataset. 
We compare our method with Cosmic Emu \citep{moran2023mira}, the state-of-the-art emulator 
constructed on the same $m=111$ Mira-Titan cosmologies. Cosmic Emu uses a Bayesian approach 
where the ``true'' spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep[DPC,][]{moran2023mira}. Cosmic Emu also uses a PC approach to predict power matter 
spectra for held-out cosmologies as a function of $\psi$ \citep{gattiker2020sepia}.

Using our proposed DGP.FCO approach and Cosmic Emu, we obtain two sets of predictions for $\hat{S}_t(\psi_t)$ 
with $t\in\{1,\ldots,6\}$ indexing the six held-out cosmologies.  In this real-world example, we have
no ``truth" for the Mira-Titan hold-outs against which to benchmark. In an effort to assess the 
accuracy of these predictions, we also separately fit the DGP.FCO model on each of these hold outs
and use the corresponding posterior means as a reference. The individual predictions for each method and 
cosmology are shown in Figure \ref{fig:plot_pred_1to6}, where the posterior mean is subtracted so that
predictions lines closer to zero represent a better fit.

\begin{figure}[!b]
    \centering
    \includegraphics[width=\textwidth]{pred_1to6.jpeg}
    \caption{Results of all six predictions for both methods on the held-out 
    cosmologies (DGP.FCO posterior mean subtracted). CosmicEMU is dotted blue and DGP.FCO is dashed orange.
    Closer to the dotted zero line indicates a better fit with the DGP.FCO posterior mean.}
    \label{fig:plot_pred_1to6}
\end{figure}

From these predictions, we can calculate MSE when treating the trained DGP.FCO predictions as a benchmark. 
Our DGP.FCO prediction achieves the greatest reduction in MSE in the region where perturbation theory
and the low-resolution runs are deemed unbiased (see Figure \ref{fig:mse_by_k}). For the region where only 
the high-resolution run is unbiased, Cosmic Emu generally has a slightly lower average RMSE. Our average MSE 
is lower at 55\% of $k$ values considered. These results indicate that DGP.FCO competes favorably with the 
state-of-the-art Cosmic Emu model.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mse_by_k.jpeg}
    \caption{Average MSE across each of the 6 hold out cosmologies, for each $k$ value. Cosmic Emu 
    predictions are shown in dotted blue, and DGP.FCO predictions are shown in dashed orange.}
    \label{fig:mse_by_k}
\end{figure}

@@

In this work, we introduce a novel Bayesian hierarchical framework that leverages deep 
Gaussian processes (DGPs) to model smooth latent functions from correlated functional outputs, 
such as those arising in cosmological simulations. Our model, DGP.FCO, builds on the compositional nature 
of DGPs and introduces a third layer to explicitly model correlated observational errors within 
each spectrum. Through a simulation study, we demonstrate that DGP.FCO offers improved uncertainty 
quantification compared to existing methods.

We apply DGP.FCO to both the CAMB and Mira-Titan datasets, generating posterior mean estimates 
and predicting at held-out cosmological settings. Across both datasets, DGP.FCO 
consistently performs well against competitive baselines, showcasing strong generalization and 
flexibility in complex, functional data regimes.

Looking ahead, several avenues for future research arise. Beyond modeling dark matter power 
spectra, DGP.FCO could be extended to hydrodynamical simulation outputs in cosmology, which 
often share similar functional structures and error correlations. More broadly, this approach may be 
applicable in other scientific domains involving structured functional outputs, especially those 
with spatial or temporal dependencies.

In addition, generating predictions over large cosmological grids opens the door to detailed 
sensitivity analyses, enabling researchers to disentangle parameter interactions and identify 
the most influential physical drivers of the spectra. From a modeling standpoint, further 
developments could include joint modeling of latent layers across cosmologies, and extensions 
to redshifts beyond $z=0$. Lastly, a hybrid framework combining DGP.FCO with Cosmic Emu, leveraging 
the strengths of each at different scales of $k$, may yield even more accurate predictions.

@@
